---
title: "Forecast on Deaths (ARIMA Modelling)"
output: html_notebook
editor_options: 
  markdown: 
    wrap: sentence
---

**Data Preparation**

```{r}
#For forecasting, we chose the latest data
trend_death_hb <- trend_hb_daily %>% 
  filter (hb_name == "Scotland") %>% 
  filter(date >="2021-06-01") %>% 
  filter(!(is.na(daily_deaths))) %>% 
  select(date, daily_deaths)

# Convert it into a time series
daily_death_hb_zoo <- zoo(trend_death_hb$daily_deaths, 
           order.by=as.Date(trend_death_hb$date, format='%m/%d/%Y'))

# Convert it into a time series
daily_death_hb_timeseries <-  timeSeries::as.timeSeries(daily_death_hb_zoo)
```

## Step 1 : Visualize the time series

```{r}
original_series_death<-autoplot(daily_death_hb_timeseries, ts.colour = '#5ab4ac')+
  xlab("Month") + 
  ylab("Patient died")+
  #scale_x_date(breaks = "1 month", date_labels = "%b - %y" )+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  ggtitle("Trend on Deaths") +
  color_theme()

ggplotly(original_series_death)
```

## Step 2 : Identification of model : (Finding d:)

Identify whether the time series is stationary / non stationary we can use ADF Augmented Dickey-Fuller test

```{r}
adf_test_death <- adf.test(daily_death_hb_timeseries)
adf_test_death
```

The time series is not stationary since we have a high p-value (p-value must be \< 0.05).
So we apply difference

```{r}
first_diff_death<- diff(daily_death_hb_timeseries)
adf_test1_death <- adf.test(na.omit(first_diff_death))
adf_test1_death
```

Create a dataframe to compare

```{r}
adf_data_death <- data.frame(Data = c("Original", "First-Ordered"),
                       Dickey_Fuller = c(adf_test_death$statistic, adf_test1_death$statistic),
                       p_value = c(adf_test_death$p.value,adf_test1_death$p.value))
adf_data_death
```

Initially the p-value is high which indicates that the Time Series is not stationary.
So we apply difference 1 time.
After the first difference, the p-value \< significance level (0.05) So we can conclude that the difference data are stationary.
So difference (d = 1)

Other method:

```{r}
ndiffs(daily_death_hb_timeseries)
```

Let's plot the First Order Difference Series

Order of first difference

```{r}

first_diff_death<- diff(daily_death_hb_timeseries)
p<- autoplot(first_diff_death, ts.colour = '#5ab4ac') +
  xlab("Month") + 
  ylab("DEATH")+
 # scale_x_date(breaks = "1 month", date_labels = "%b - %y" )+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  ggtitle("First-Order Difference Series") +
  color_theme()

ggplotly(p)
```

## Step 3 Estimate the parameters (Finding p and q)

For our model ARIMA (p,d,q), we found d = 1, the next step is to get the values of p and q, the order of AR and MA part.
Plot ACF and PACF charts to identify q and p respectively.

```{r}
par(mfrow=c(2,2))
acf_death  <- acf1(first_diff_death, col=2:7, lwd=4)
pacf_death <- acf1(first_diff_death,  pacf = TRUE, col=2:7, lwd=4)
```

The ACF and PACF plots tapers to zeo at one point.
So it follows an ARMA model

The PACF non-zero value at lag 3 So the data may follow an AR(3) model

The ACF non - zero value at lag 2 So it can be MA (2)

So we propose three ARMA models for the differenced data: ARMA(p,q) ARMA(3,2), ARMA(3,0) and ARMA(0,2).

That is, for the original time series, we propose three ARIMA models,ARIMA(p,d,q) ARIMA(3,1,2), ARIMA(3,1,0) and ARMA(0,1,2).

## Step 4 Build the ARIMA model

### Manual ARIMA

```{r}
arima_fit_dth_1 = Arima(daily_death_hb_timeseries, order = c(3,1,2))
arima_fit_dth_2 = Arima(daily_death_hb_timeseries, order = c(3,1,0))
arima_fit_dth_3 = Arima(daily_death_hb_timeseries, order = c(0,1,2))
```

```{r}
summary(arima_fit_dth_1)
summary(arima_fit_dth_2)
summary(arima_fit_dth_3)
```

Another way of checking AIC

```{r}
texreg::screenreg(list(arima_fit_dth_1, arima_fit_dth_2, arima_fit_dth_3),
                custom.model.names =c("ARIMA(3,1,2)","ARIMA(3,1,0)","ARIMA(0,1,2)"),
                center = TRUE,
                table = FALSE)
```

Based on this, ARIMA model (3,1,2) seems best.
We can verify the same using automated method

### Automated ARIMA

```{r}
auto_arima_fit_death <- auto.arima(lag(daily_death_hb_timeseries),
                  seasonal=FALSE,
                  stepwise=FALSE,
                  approximation=FALSE,
                  trace = TRUE
                  )
auto_arima_fit_death
```

However Automated ARIMA also confirms that the ARIMA(2,1,3) seems good based on AIC

```{r}
coef_dth<-lmtest::coeftest(auto_arima_fit_death)
coef_dth
```

**Model Selection Criteria :**

ARIMA models with minimum AIC, RMSE and MAPE criteria were chosen as the best models.
Based on Akaike Information Criterion (AIC) above, an ARIMA(2, 1, 3) model seems best.

## Step 5 Check for Diagnostics

Let's plot the diagnostics with the results to make sure the normality and correlation assumptions for the model hold.
If the residuals look like white noise, proceed with forecast and prediction, otherwise repeat the model building.

```{r}
res_dth <-checkresiduals(auto_arima_fit_death, theme = color_theme())
res_dth
```

The ACF plot of the residuals from the ARIMA(2,1,3) model shows that all auto correlations are within the threshold limits except 1, But portmanteau test (Ljung-Box test) shows a higher p-value which means the values are indpendent.
i.e We fail to reject the null hypothesis

Fitting the ARIMA model with the existing data

Let's plot the actuals against the fitted values

**Convert model and time series to dataframe for plotting**

```{r}
daily_death_hb_timeseries_data <- fortify(daily_death_hb_timeseries) %>% 
  clean_names() %>% 
  remove_rownames %>% 
  rename (date = index,
          death = data)%>% 
  mutate(index = seq(1:nrow(daily_death_hb_timeseries)))
  
arima_fit_dth_resid <- ts(daily_death_hb_timeseries[1:nrow(daily_death_hb_timeseries)]) - resid(auto_arima_fit_death)

arima_fit_dth_data <- fortify(arima_fit_dth_resid) %>% 
  clean_names() %>% 
  mutate(data = round(data,2))%>% 
  mutate (data = ifelse(data < 0,0,data))

fit_existing_dth_data <- daily_death_hb_timeseries_data %>% 
  inner_join(arima_fit_dth_data, by = c("index"))


```

**Plotting the series along with the fitted values**

```{r}

fit_existing_dth_plot <- fit_existing_dth_data %>% 
   mutate (data = ifelse(data < 0,0,data)) %>% 
  mutate(date = as.Date(date)) %>% 
  ggplot()+
  aes(x=date, y = death)+
  geom_line(color ="#5ab4ac")+
  geom_line(aes(x= date, y = data), colour = "red" )+
  xlab("Month") + 
  ylab("Deaths reported")+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  scale_x_date(breaks = "1 month", date_labels = "%b - %y" )+
  ggtitle("Fitting the ARIMA model with existing data") +
  color_theme()

ggplotly(fit_existing_dth_plot)

```

## Step 6 Forecast using the model

**Data Preparation :**

```{r}
forecast_dth_model <- forecast(auto_arima_fit_death,level = c(80, 95), h = 30) 

#Convert the model to dataframe for plotting

# Negative values of the CI interval are considered as 0

forecast_dth_model_data <- fortify(forecast_dth_model) %>% 
  clean_names() %>% 
  mutate(data = round(data,2),
         fitted= round(fitted,2))  %>% 
  mutate (lo_80 = ifelse(lo_80 < 0,0,lo_80),
          lo_95 = ifelse(lo_95 < 0,0,lo_95)
          )

forecast_start_date <- as.Date(max(daily_death_hb_timeseries_data$date)+1)
forecast_end_date <- as.Date(forecast_start_date+29)

forecast_dth_data <- forecast_dth_model_data %>% 
  filter(!(is.na(point_forecast))) %>% 
  mutate(date = seq(forecast_start_date,forecast_end_date, by =1)) %>% 
select(-data,-fitted, -index)  

fitted_dth_data <- forecast_dth_model_data %>% 
  filter(!(is.na(data))) %>% 
  inner_join(daily_death_hb_timeseries_data, by = c("index")) %>% 
  mutate(date = as.Date(date)) %>% 
select(date, data, fitted) 

```

**Plotting the Vaccination series plus the forecast and 80 - 95% prediction intervals**

```{r}

annotation <- data.frame(
   x = c(as.Date("13-06-2021","%d-%m-%Y"),as.Date("11-10-2021","%d-%m-%Y")),
   y = c(30,40),
   label = c("PAST", "FUTURE")
)

#Time series plots for the next 60 days according to best ARIMA models with 80%–95% CI.
forecast_data_dth_plot <- fitted_dth_data %>% 
  ggplot()+
  geom_line(aes(x= date, y = data), color = "#5ab4ac")+
  #geom_line(aes(x= date, y = fitted), colour = "red" )+
  geom_line(aes(x= date, y =point_forecast), color ="blue", data = forecast_dth_data )+
  geom_ribbon(aes(x = date, y = point_forecast, ymin = lo_80, ymax = hi_80), 
              data = forecast_dth_data, alpha = 0.3, fill = "green")+
  geom_ribbon(aes(x = date, y = point_forecast, ymin = lo_95, ymax = hi_95), 
              data = forecast_dth_data, alpha = 0.1)+
  geom_vline(aes(xintercept=as.numeric(max(date))),color="#f1a340", linetype="dashed",data = fitted_dth_data)+
  ggtitle("Projection of new Deaths") +
  xlab("Month") + 
  ylab("Death reported")+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  color_theme()+
  scale_x_date(breaks = "1 month", date_labels = "%b - %y" )+
  geom_text(data=annotation, 
            aes( x=x, y=y, label=label),                  
            color="blue", 
            size=4 )
   

ggplotly(forecast_data_dth_plot )
```
