---
title: "PHS_COVID_ALL"
output: html_notebook
---

```{r message=FALSE, warning=FALSE, include=FALSE}
#Load the data------------------------------------------------------------------
library(here)
library(ggplot2)
library(plotly)
library(fable)
library(fabletools)
library(tsibble)
library(RColorBrewer)
library(leaflet)
library(tseries)
library(forecast)
library(zoo)
library(dlm)
library(urca)
library(ggfortify)
library(gridExtra)
library(astsa)
```

```{r message=FALSE, warning=FALSE, include=FALSE}

#Load the data------------------------------------------------------------------
source(here("cleaning_scripts/phs_covid_cleaning.R"))
source(here("analysis/phs_covid_functions.R"))
```

### ***Plot1(a): Trend on people who tested positive***

```{r}
#Calculate rolling average for Health board equals Scotland   
trend_hb_daily_roll_avg <- calculate_roll_avg("positive")

#convert the data frame to time series to use the range slider and selector
trend_hb_daily_roll_avg_ts <- convert_to_timeseries(trend_hb_daily_roll_avg)

# Create the plot using plot_ly
trend_positive_plot <- plot_ly(trend_hb_daily_roll_avg_ts, x = ~Date,
                          hoverinfo ="text",
                          text = ~paste("Date: ", Date,"<br>",
                                        "Daily Positive: ", daily_positive,"<br>",
                                        "Seven Day Average: ",seven_day_roll_avg))  
#add bar chart for daily positive
trend_positive_plot <- trend_positive_plot %>% add_bars(y = ~daily_positive, 
                                              name = "Daily Positive",
                                              color  =  I("#5ab4ac")  
                                              )
#add line for rolling average
trend_positive_plot <- trend_positive_plot %>% add_lines(y = ~seven_day_roll_avg,
                                               name = "Seven Day Rolling Average",
                                               color = I("red")) 
#add layout to the existing plot
trend_positive_plot <- trend_positive_plot %>% layout(
    title = list(text ="Trend on Positive Cases"),
    xaxis = xaxis_list,
    yaxis = yaxis_list,
    legend = list(title=list(text='<b> Trend </b>'),
                  x = 0.1, y = 0.9))
     
trend_positive_plot
```

Data for the report (Presentation)

```{r}
total_positive_case <- trend_hb_daily %>% 
  filter(hb_name == "Scotland") %>% 
  filter(date == max(date))

recent_7days <- calculate_roll_avg("positive") %>% 
  filter (date >= max(date)-7) %>% 
  summarise(total = sum(daily_positive))

previous_7days <- calculate_roll_avg("positive") %>% 
  filter (date >= max(date)-14) %>% 
  filter (date < max(date)-7) %>% 
  summarise(total = sum(daily_positive))

percentage_calc = (recent_7days -previous_7days )/previous_7days * 100 

percentage_calc
# decrease of 13.04 %

total_hospital <- trend_hb_daily %>% 
  filter(hb_name == "Scotland") %>% 
  summarise(total = sum(hospital_admissions, na.rm = TRUE))

recent_7days_hospital <- calculate_roll_avg("hospital") %>% 
  filter (!(is.na(seven_day_roll_avg))) %>% 
  filter (date >= max(date)-7) %>% 
  summarise(total = sum(hospital_admissions))

previous_7days_hospital <- calculate_roll_avg("hospital") %>% 
  filter (!(is.na(seven_day_roll_avg))) %>% 
  filter (date >= max(date)-14) %>% 
  filter (date < max(date)-7) %>% 
  summarise(total = sum(hospital_admissions))

percentage_calc_hosp = (recent_7days_hospital -previous_7days_hospital )/previous_7days_hospital * 100 

percentage_calc_hosp
# decrease of 15.11 %


icu_hospital <- trend_hb_daily %>% 
  filter(hb_name == "Scotland") %>% 
  summarise(total = sum(icu_admissions, na.rm = TRUE))

cum_deaths <- trend_hb_daily %>% 
  filter(hb_name == "Scotland") %>% 
  filter (date == max(date)) %>% 
  select (cumulative_deaths)

daily_vacc_hb %>% 
  filter(hb_name == "Scotland",
         sex == "Total",
         age_group == "All vaccinations") %>% 
  filter (dose == "Dose 1") %>% 
  filter (date == max(date))
```

### ***Plot1(b): Total Positive Case by neighborhood (Based on Local Authority).***

```{r}

trend_la_daily_nb <- calculate_roll_avg("la_neighbour")

# Get the population data  
la_population <-trend_seven_day %>% 
  filter(date == max(date)) %>% 
  group_by (ca) %>% 
  summarise(total_population = sum(population))

# Prepare the map data by combining spatial data and (local authority) la data  
map_data <- zones_la %>% 
    inner_join(trend_la_daily_nb, by = c("code" = "ca"))%>% 
    inner_join(la_population, by = c("code" = "ca")) %>% 
  arrange(seven_day_roll_avg)

# Create a map palette  
map_palette <- colorNumeric("plasma", domain = range(map_data$seven_day_roll_avg))
map_data <- map_data %>% 
    mutate(colour = map_palette(seven_day_roll_avg))

# Display the spatial data
map_data %>%
    leaflet() %>%
    addPolygons(
     popup = ~ str_c("<b><h2>",name,  "</h2>", 
                      "<h3>(01-Oct-2011 to 07-Oct-2021)</h3></b>", 
                      "<b>Number of Positive Cases over 7 days: </b>", seven_day_roll_avg,
                     " <br><b>7 day rate per 100,000 people: </b>", crude_rate7day_positive,
                     " <br><br><b>Total Population: </b>", total_population,
                     
                      sep = ""),
      color = ~colour
    ) %>%
    addLegend(
      position = "topright",
      colors = ~colour,
      labels = ~name
    )

```

### **2 Analyse the trend on Hospitalizations:**

```{r}
#Calculate rolling average for Health board equals Scotland   
trend_hb_hosp_roll_avg <- calculate_roll_avg("hospital")

#convert the data frame to time series to use the range slider and selector
trend_hb_hosp_roll_avg_ts <- convert_to_timeseries(trend_hb_hosp_roll_avg)

# Create the plot using plot_ly
trend_hospital_plot <- plot_ly(trend_hb_hosp_roll_avg_ts, x = ~Date,
                          hoverinfo ="text",
                          text = ~paste("Date: ", Date,"<br>",
                                        "Patients admitted: ", hospital_admissions,"<br>",
                                        "Seven Day Average: ",seven_day_roll_avg))  
#add bar chart for daily positive
trend_hospital_plot <- trend_hospital_plot %>% add_bars(y = ~hospital_admissions, 
                                              name = "Patients admitted",
                                              color  =  I("#5ab4ac")  
                                              )
#add line for rolling average
trend_hospital_plot <- trend_hospital_plot %>% add_lines(y = ~seven_day_roll_avg,
                                               name = "Seven Day Rolling Average",
                                               color = I("red")) 
#add layout to the existing plot
trend_hospital_plot <- trend_hospital_plot %>% layout(
    title = list(text ="Hospitalisation"),
    xaxis = xaxis_list,
    yaxis = yaxis_list,
    legend = list(title=list(text='<b> Trend </b>'),
                  x = 0.1, y = 0.9))
     
trend_hospital_plot

```

### **2 Analyse the trend on Deaths:**

```{r}
#Calculate rolling average for Health board equals Scotland   
trend_hb_death_roll_avg <- calculate_roll_avg("death")

#convert the data frame to time series to use the range slider and selector
trend_hb_death_roll_avg_ts <- convert_to_timeseries(trend_hb_death_roll_avg)

# Create the plot using plot_ly
trend_death_plot <- plot_ly(trend_hb_death_roll_avg_ts, x = ~Date,
                          hoverinfo ="text",
                          text = ~paste("Date: ", Date,"<br>",
                                        "Daily Deaths: ", daily_deaths,"<br>",
                                        "Seven Day Average: ",seven_day_roll_avg))  
#add bar chart for daily positive
trend_death_plot <- trend_death_plot %>% add_bars(y = ~daily_deaths, 
                                              name = "Daily Deaths",
                                              color  =  I("#5ab4ac")  
                                              )
#add line for rolling average
trend_death_plot <- trend_death_plot %>% add_lines(y = ~seven_day_roll_avg,
                                               name = "Seven Day Rolling Average",
                                               color = I("red")) 
#add layout to the existing plot
trend_death_plot <- trend_death_plot %>% layout(
    title = list(text ="Deaths reported"),
    xaxis = xaxis_list,
    yaxis = yaxis_list,
    legend = list(title=list(text='<b> Trend </b>'),
                  x = 0.1, y = 0.9))
     
trend_death_plot

```

"PHS_COVID Vaccination Prediction Model Using ARIMA"

### Data Preparation

```{r}

trend_vacc_hb <- daily_vacc_hb %>% 
  filter(hb_name == "Scotland") %>% 
  filter(sex =="Total") %>% 
  filter(age_group == "All vaccinations") %>% 
  filter(cumulative_number_vaccinated!=0) 

```

## **Analyse the trend on Vaccinations:**

### ***Plot1(a): Trend on Vaccination***

```{r}
#Plot to visualize trend on vaccination.
plot_vaccine <- trend_vacc_hb %>% 
  ggplot()+
  aes(x = date, y = number_vaccinated)+
  geom_line(aes(color = dose))+
  scale_x_date(breaks = "1 month", date_labels = "%b - %y" )+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  ggtitle("People Vaccinated") +
  xlab("Year") +
  ylab("No of Positive Cases") +
  color_theme()+
  scale_colour_manual(values = c("#f1a340", "#5ab4ac"))

ggplotly(plot_vaccine)
```

### ***Plot1(b): Cumulative Total on Vaccination***

```{r}
# Identify the population
dose_population <- daily_vacc_hb %>% 
  filter(sex == "Total") %>% 
  filter(date == max(date)) %>% 
  filter(hb_name =="Scotland") %>% 
  filter(age_group =="16 years and over") %>% 
  select(population) %>% 
  distinct()

#Plot to visualise cumulative vaccination trend.  
plot_vaccine_cumm <- trend_vacc_hb %>% 
  ggplot()+
  aes(x = date, y = cumulative_number_vaccinated)+
  geom_line(aes(color = dose))+
  scale_x_date(breaks = "1 month", date_labels = "%b - %y" )+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  ggtitle("Cummulative Trend on Vaccination") +
  xlab("Year") +
  ylab("No of People Vaccinated") +
  color_theme()+
  scale_colour_manual(values = c("#f1a340", "#5ab4ac"))+
  scale_y_continuous(labels = scales::unit_format(unit = "M", scale = 1e-6), 
                     sec.axis = sec_axis(trans = ~./dose_population$population,
                     name = "Percentage",
                     labels = scales::label_percent(accuracy = 0.01)
                    ))

plotly::ggplotly(plot_vaccine_cumm)

```

### ***Plot1(c): Cumulative Total on Vaccination based on Age group***

```{r}
cumm_vac_age<- daily_vacc_hb %>% 
  filter(sex == "Total") %>% 
  filter(is.na(age_group_qf)) %>% 
  filter(date == max(date)) %>% 
  filter(hb_name =="Scotland") %>% 
  filter(age_group !="All vaccinations") %>% 
  mutate (cumulative_percent_coverage = ifelse(cumulative_percent_coverage >100, 100,
                                               round(cumulative_percent_coverage,2))) %>% 
  select(dose,age_group, cumulative_percent_coverage, population)
  
cumm_vac_age_plot <- cumm_vac_age %>% 
  ggplot()+
  aes(x = age_group, y = cumulative_percent_coverage)+
  geom_col(aes(fill = dose), position = "dodge")+
   # geom_text(aes(label=cumulative_percent_coverage, hjust = 0),
   #           position=position_dodge(width=0.9),angle = 90)+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  ggtitle("Percentage Coverage of Vaccination by age group") +
  xlab("Age group") +
  ylab("% of Coverage") +
  color_theme()+
  scale_fill_manual(values = c("#f1a340", "#5ab4ac"))

ggplotly(cumm_vac_age_plot)

#position=position_dodge(width=0.9), vjust=-0.25

```

# Forecast on Vaccination: ARIMA Model

**Data Preparation**

```{r}
trend_vacc_hb <- trend_vacc_hb %>% 
  filter (dose == "Dose 2") %>% 
  select(date,cumulative_number_vaccinated)

# Convert it to zoo type
daily_vacc_hb_zoo <- zoo(trend_vacc_hb$cumulative_number_vaccinated, 
           order.by=as.Date(trend_vacc_hb$date, format='%m/%d/%Y'))

# Convert it into a time series
daily_vacc_hb_timeseries <-timeSeries::as.timeSeries(daily_vacc_hb_zoo)

```

## Step 1 : Visualize the time series

```{r}
original_series<-
  autoplot(daily_vacc_hb_timeseries, colour = '#5ab4ac')+
  xlab("Month") + 
  ylab("VACCINATED")+
  #scale_x_date(breaks = "1 month", date_labels = "%b - %y" )+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  ggtitle("Original Series") +
  scale_y_continuous(labels = scales::unit_format(unit = "M", scale = 1e-6))+
  color_theme()

ggplotly(original_series)
```

## Step 2 : Identification of model : (Finding d:)

Identify whether the time series is stationary / non stationary we can use ADF Augmented Dickey-Fuller test

```{r}
adf_test <- adf.test(daily_vacc_hb_timeseries)
```

The time series is not stationary since we have a high p-value.
So we apply difference

```{r}
first_diff_ts<- diff(daily_vacc_hb_timeseries)
adf_test1 <- adf.test(na.omit(first_diff_ts))
second_diff_ts<- diff(first_diff_ts)
adf_test2 <- adf.test(na.omit(second_diff_ts))

adf_test1
adf_test2
```

Create a dataframe to compare

```{r}
adf_data <- data.frame(Data = c("Original", "First-Ordered", "Second Ordered"),
                       Dickey_Fuller = c(adf_test$statistic, adf_test1$statistic, adf_test2$statistic),
                       p_value = c(adf_test$p.value,adf_test1$p.value,adf_test2$p.value))
adf_data
```

Initially the p-value is high which indicates that the Time Series is not stationary.
So we apply difference 2 times.
After the second difference, the p-value \< significance level (0.05) So we can conclude that the difference data are stationary.
So difference (d = 2)

Other method to confirm

```{r}
ndiffs(daily_vacc_hb_timeseries)
```

Let's plot the First Order and Second Order Difference Series

Order of first difference

```{r}

first_order<- autoplot(first_diff_ts, ts.colour = '#5ab4ac') +
  xlab("Month") + 
  ylab("VACCINATED")+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  ggtitle("First-Order Difference") +
  color_theme()

ggplotly(first_order)
```

Order of Second difference

```{r}

second_order<- autoplot(second_diff_ts, ts.colour = '#5ab4ac') +
  xlab("Month") + 
  ylab("VACCINATED")+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  ggtitle("Second-Order Difference") +
  color_theme()

ggplotly(second_order)
```

## Step 3 Estimate the parameters (Finding p and q)

For our model ARIMA (p,d,q), we found d = 2, the next step is to get the values of p and q, the order of AR and MA part.
Plot ACF and PACF charts to identify q and p respectively.

```{r echo=FALSE, message=FALSE, warning=FALSE}
par(mfrow=c(2,2))
  acf1(first_diff_ts, col=2:7, lwd=4)
  acf1(second_diff_ts, col=2:7, lwd=4)
  acf1(first_diff_ts, pacf = TRUE, col=2:7, lwd=4)
  acf1(second_diff_ts, pacf = TRUE, col=2:7, lwd=4)
```

The ACF and PACF plots of the differenced data show the following patterns:

The ACF doesn't follow a sinusoidal pattern but its slowly geometric decay.
Also there is a significant spike at lag 3 in the PACF, but none beyond lag 3.
So the data may follow an AR(3) model

The PACF is sinusoidal and decaying.
Also there is a significant spike at lag 2 in the ACF, but none beyond lag 2 So the data may follow an MA(2) model

So we propose three ARMA models for the differenced data: ARMA(p,q) ARMA(3,2), ARMA(3,0) and ARMA(0,2).

That is, for the original time series, we propose three ARIMA models,ARIMA(p,d,q) ARIMA(3,1,2), ARIMA(3,1,0) and ARMA(3,1,2).

## Step 4 Build the ARIMA model

### Manual ARIMA:

```{r}
arima_fit1 = Arima(daily_vacc_hb_timeseries, order = c(3,1,2))
arima_fit2 = Arima(daily_vacc_hb_timeseries, order = c(3,1,0))
arima_fit3 = Arima(daily_vacc_hb_timeseries, order = c(3,1,2))
arima_fit4 = Arima(daily_vacc_hb_timeseries, order = c(3,1,1))
```

```{r}
summary(arima_fit1)
summary(arima_fit2)
summary(arima_fit3)
summary(arima_fit4)
```

Forecast the Manual ARIMA model

```{r}
# Forecast the manual models

future = forecast(arima_fit1, h = 30)
future2 = forecast(arima_fit2, h = 30)
future3 = forecast(arima_fit3, h = 30)
future4 = forecast(arima_fit4, h = 30)

#Plot the forecasted manual models

par(mfrow = c(2,2))
plot(future)
plot(future2)
plot(future3)
plot(future4)
```

### **Automated ARIMA**

```{r}
auto_arima_fit_vacc <- auto.arima(daily_vacc_hb_timeseries,
                  seasonal=FALSE,
                  stepwise = FALSE,
                  approximation = FALSE,
                  trace = TRUE
                  )
summary(auto_arima_fit_vacc)
```

**Model Selection Criteria :**

ARIMA models with minimum AIC, RMSE and MAPE criteria were chosen as the best models.
Automated ARIMA confirms that the ARIMA(3, 2, 2) seems good based on AIC

```{r}
lmtest::coeftest(auto_arima_fit_vacc)
```

All the coefficients are statistically significant.

## Step 5 Check for Diagnostics

Let's plot the diagnostics with the results to make sure the normality and correlation assumptions for the model hold.
If the residuals look like white noise, proceed with forecast and prediction, otherwise repeat the model building.

```{r}
res <- checkresiduals(auto_arima_fit_vacc, theme = color_theme())
res
```

The ACF plot of the residuals from the ARIMA(3,2,2) model shows that almost auto correlationswith regular interval outlier.
A portmanteau test returns a smaller p-value (almost close to Zero), also suggesting that the residuals are white noise.

**Fitting the ARIMA model with the existing data**

The residual errors seem fine with near zero mean and uniform variance.
Let's plot the actuals against the fitted values

```{r}
#Convert the model to dataframe for plotting

daily_vacc_hb_timeseries_data <- fortify(daily_vacc_hb_timeseries) %>% 
  clean_names() %>% 
  remove_rownames %>% 
  rename (date = index,
          vacc = data)%>% 
  mutate(index = seq(1:nrow(daily_vacc_hb_timeseries)))
  
arima_fit_resid <- ts(daily_vacc_hb_timeseries) - resid(auto_arima_fit_vacc)

arima_fit_data <- fortify(arima_fit_resid) %>% 
  clean_names() %>% 
  mutate(data = round(data,2))

fit_existing_data <- daily_vacc_hb_timeseries_data %>% 
  inner_join(arima_fit_data, by = c("index"))
```

```{r}
#plotting the series along with the fitted values
fit_existing_data %>% 
  ggplot()+
  aes(x=date, y = vacc)+
  geom_line(color ="#5ab4ac")+
  geom_line(aes(x= date, y = data), colour = "red" )+
  xlab("Month") + 
  ylab("Number of People vaccinated")+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  ggtitle("Fitting the ARIMA model with existing data") +
  scale_y_continuous(labels = scales::unit_format(unit = "M", scale = 1e-6))+
  color_theme()
```

## Step 6 Forecast using the model

**Data Preparation :**

```{r}
#Convert the model to dataframe for plotting
forecast_model <- forecast(auto_arima_fit_vacc,level = c(80, 95), h = 60) 

forecast_model_data <- fortify(forecast_model) %>% 
  clean_names() %>% 
  mutate(data = round(data,2),
         fitted= round(fitted,2)) 

forecast_start_date <- as.Date(max(daily_vacc_hb_timeseries_data$date)+1)
forecast_end_date <- as.Date(forecast_start_date+59)

forecast_data <- forecast_model_data %>% 
  filter(!(is.na(point_forecast))) %>% 
  mutate(date = seq(forecast_start_date,forecast_end_date, by =1)) %>% 
select(-data,-fitted, -index)  

fitted_data <- forecast_model_data %>% 
  filter(!(is.na(data))) %>% 
  inner_join(daily_vacc_hb_timeseries_data, by = c("index")) %>% 
  mutate(date = as.Date(date)) %>% 
select(date, data, fitted) 

```

**Plotting the Vaccination series plus the forecast and 95% prediction intervals**

```{r}
annotation <- data.frame(
   x = c(as.Date("03-04-2021","%d-%m-%Y"),as.Date("31-10-2021","%d-%m-%Y")),
   y = c(1000000,3000000),
   label = c("PAST", "FUTURE")
)

#Time series plots for the next 60 days according to best ARIMA models with 80%–95% CI.
fitted_data %>% 
  ggplot()+
  geom_line(aes(x= date, y = data))+
  geom_line(aes(x= date, y = fitted), colour = "red" )+
  geom_line(aes(x= date, y =point_forecast), data = forecast_data )+
  geom_ribbon(aes(x = date, y = point_forecast, ymin = lo_80, ymax = hi_80), 
              data = forecast_data, alpha = 0.3, fill = "green")+
  geom_ribbon(aes(x = date, y = point_forecast, ymin = lo_95, ymax = hi_95), 
              data = forecast_data, alpha = 0.1)+
  ggtitle("Forecast") +
  xlab("Month") + 
  ylab("Number of People vaccinated")+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  color_theme()+
  scale_y_continuous(labels = scales::unit_format(unit = "M", scale = 1e-6))+
  scale_x_date(breaks = "1 month", date_labels = "%b - %y" )+
   geom_text(data=annotation, 
             aes( x=x, y=y, label=label),                  
            color="red", 
            size=4 )+
  geom_vline(xintercept =as.Date("08-10-2021","%d-%m-%Y"), linetype = 2)
```

"Forecast on Hospitalisation (ARIMA Modelling)"

**Data Preparation**

```{r}
#For forecasting, we chose the latest data
trend_hosp_hb <- trend_hb_daily %>% 
  filter (hb_name == "Scotland") %>% 
  filter(date >="2021-06-01") %>% 
  filter(!(is.na(hospital_admissions))) %>% 
  select(date, hospital_admissions)

# Convert it into a time series
daily_hosp_hb_zoo <- zoo(trend_hosp_hb$hospital_admissions, 
           order.by=as.Date(trend_hosp_hb$date, format='%m/%d/%Y'))

# Convert it into a time series
daily_hosp_hb_timeseries <-  timeSeries::as.timeSeries(daily_hosp_hb_zoo)
```

## Step 1 : Visualize the time series

```{r}
original_series<-autoplot(daily_hosp_hb_timeseries, ts.colour = '#5ab4ac')+
  xlab("Month") + 
  ylab("Number of People hospitalised")+
  #scale_x_date(breaks = "1 month", date_labels = "%b - %y" )+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  ggtitle("Trend on Hospitalisation") +
  color_theme()

ggplotly(original_series)
```

## Step 2 : Identification of model : (Finding d:)

Identify whether the time series is stationary / non stationary we can use ADF Augmented Dickey-Fuller test

```{r}
adf_test_hosp <- adf.test(daily_hosp_hb_timeseries)
adf_test_hosp
```

The time series is not stationary since we have a high p-value (p-value must be \< 0.05).
So we apply difference

```{r}
first_diff_hosp<- diff(daily_hosp_hb_timeseries)
adf_test1_hosp <- adf.test(na.omit(first_diff_hosp))
adf_test1_hosp
```

Create a dataframe to compare

```{r}
adf_data_hosp <- data.frame(Data = c("Original", "First-Ordered"),
                       Dickey_Fuller = c(adf_test_hosp$statistic, adf_test1_hosp$statistic),
                       p_value = c(adf_test_hosp$p.value,adf_test1_hosp$p.value))
adf_data_hosp
```

Initially the p-value is high which indicates that the Time Series is not stationary.
So we apply difference 1 time.
After the first difference, the p-value \< significance level (0.05) So we can conclude that the difference data are stationary.
So difference (d = 1)

Other method:

```{r}
ndiffs(daily_hosp_hb_timeseries)
```

Let's plot the First Order Difference Series

Order of first difference

```{r}
p<- autoplot(first_diff_hosp, ts.colour = '#5ab4ac') +
  xlab("Month") + 
  ylab("HOSPITALIZATION")+
 # scale_x_date(breaks = "1 month", date_labels = "%b - %y" )+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  ggtitle("First-Order Difference Series") +
  color_theme()

ggplotly(p)
```

## Step 3: Estimate the parameters (Finding p and q)

For our model ARIMA (p,d,q), we found d = 1, the next step is to get the values of p and q, the order of AR and MA part.
Plot ACF and PACF charts to identify q and p respectively.

```{r}
par(mfrow=c(2,2))
acf_hosp  <- acf1(daily_hosp_hb_timeseries, col=2:7, lwd=4)
pacf_hosp <- acf1(daily_hosp_hb_timeseries,  pacf = TRUE, col=2:7, lwd=4)
acf_hosp  <- acf1(first_diff_hosp, col=2:7, lwd=4)
pacf_hosp <- acf1(first_diff_hosp,  pacf = TRUE, col=2:7, lwd=4)
```

The ACF and PACF plots of the differenced data show the following patterns:

The ACF is sinusoidal and there is a significant spike at lag 3 in the PACF So the data may follow an AR(3) model

The PACF is sinusoidal and there is a significant spike at lag 2 in the ACF So the data may follow an MA(2) model

So we propose three ARMA models for the differenced data: ARMA(p,q) ARMA(3,2), ARMA(3,0) and ARMA(0,2).

That is, for the original time series, we propose three ARIMA models,ARIMA(p,d,q) ARIMA(3,1,2), ARIMA(3,1,0) and ARMA(0,1,2).

## Step 4: Build the ARIMA model

### **Manual ARIMA:**

```{r}
arima_fit_hosp_1 = Arima(daily_hosp_hb_timeseries, order = c(3,1,2))
arima_fit_hosp_2 = Arima(daily_hosp_hb_timeseries, order = c(3,1,0))
arima_fit_hosp_3 = Arima(daily_hosp_hb_timeseries, order = c(0,1,2))
```

```{r}
summary(arima_fit_hosp_1)
summary(arima_fit_hosp_2)
summary(arima_fit_hosp_3)
```

```{r}

texreg::screenreg(list(arima_fit_hosp_1, arima_fit_hosp_2, arima_fit_hosp_3),
                custom.model.names =c("ARIMA(3,1,2)","ARIMA(3,1,0)","ARIMA(0,1,4)"),
                center = TRUE,
                table = FALSE)
```

```{r}
#Function for Automated ARIMA


auto_arima_fit_hosp <- auto.arima(lag(daily_hosp_hb_timeseries,k =lag_value ),
                  seasonal=FALSE,
                  stepwise=FALSE,
                  approximation=FALSE,
                  trace = TRUE
                  )
```

### Automated ARIMA

```{r}
#Lag is used to best fit the model
auto_arima_fit_hosp <- auto.arima(lag(daily_hosp_hb_timeseries),
                  seasonal=FALSE,
                  stepwise=FALSE,
                  approximation=FALSE,
                  trace = TRUE
                  )
auto_arima_fit_hosp
```

Automated ARIMA confirms that the ARIMA(3,1,2) seems good based on AIC

```{r}
coef<-lmtest::coeftest(auto_arima_fit_hosp)
coef
```

All coefficients are significant except ar3.

### Model Selection Criteria :

ARIMA models with minimum AIC, RMSE and MAPE criteria were chosen as the best models.
Based on Akaike Information Criterion (AIC) above, an ARIMA(3, 1, 2) model seems best.

## Step 5: Check for Diagnostics

Let's plot the diagnostics with the results to make sure the normality and correlation assumptions for the model hold.
If the residuals look like white noise, proceed with forecast and prediction, otherwise repeat the model building.

```{r}
res <-checkresiduals(auto_arima_fit_hosp, theme = color_theme())
res
```

The ACF plot of the residuals from the ARIMA(3,1,2) model shows that all auto correlations are almost within the threshold limits, with residuals.
A portmanteau test (Ljung-Box test) returns a smaller p-value , also suggesting that the residuals are white noise.

Fitting the ARIMA model with the existing data

The residual errors seem fine with near zero mean and uniform variance.
Let's plot the actuals against the fitted values

**Convert model and time series to dataframe for plotting**

```{r}
daily_hosp_hb_timeseries_data <- fortify(daily_hosp_hb_timeseries) %>% 
  clean_names() %>% 
  remove_rownames %>% 
  rename (date = index,
          hosp = data)%>% 
  mutate(index = seq(1:nrow(daily_hosp_hb_timeseries)))
  
arima_fit_resid <- ts(daily_hosp_hb_timeseries[1:nrow(daily_hosp_hb_timeseries)]) - resid(auto_arima_fit_hosp)

arima_fit_data <- fortify(arima_fit_resid) %>% 
  clean_names() %>% 
  mutate(data = round(data,2))

fit_existing_data <- daily_hosp_hb_timeseries_data %>% 
  inner_join(arima_fit_data, by = c("index"))
```

**Plotting the series along with the fitted values**

```{r}
fit_existing_hosp_plot <- fit_existing_data %>% 
  mutate (date = as.Date(date)) %>% 
  ggplot()+
  aes(x=date, y = hosp)+
  geom_line(color ="#5ab4ac")+
  geom_line(aes(x= date, y = data), colour = "red" )+
  xlab("Month") + 
  ylab("Patient Hospitalised")+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  ggtitle("Fitting the ARIMA model with existing data") +
  scale_x_date(breaks = "1 month", date_labels = "%b - %y" )+
  #scale_y_continuous(labels = scales::unit_format(unit = "M", scale = 1e-6))+
  color_theme()

ggplotly(fit_existing_hosp_plot)
```

## Step 6 Forecast using the model

**Data Preparation :**

```{r}
forecast_model <- forecast(auto_arima_fit_hosp,level = c(80, 95), h = 30) 

#Convert the model to dataframe for plotting

forecast_model_data <- fortify(forecast_model) %>% 
  clean_names() %>% 
  mutate(data = round(data,2),
         fitted= round(fitted,2)) %>% 
  mutate (lo_80 = ifelse(lo_80 < 0,0,lo_80),
          lo_95 = ifelse(lo_95 < 0,0,lo_95)
  )

forecast_start_date <- as.Date(max(daily_hosp_hb_timeseries_data$date)+1)
forecast_end_date <- as.Date(forecast_start_date+29)

forecast_data <- forecast_model_data %>% 
  filter(!(is.na(point_forecast))) %>% 
  mutate(date = seq(forecast_start_date,forecast_end_date, by =1)) %>% 
select(-data,-fitted, -index)  

fitted_data <- forecast_model_data %>% 
  filter(!(is.na(data))) %>% 
  inner_join(daily_hosp_hb_timeseries_data, by = c("index")) %>% 
  mutate(date = as.Date(date)) %>% 
select(date, data, fitted) 

```

**Plotting the Vaccination series plus the forecast and 80 - 95% prediction intervals**

```{r}

annotation <- data.frame(
   x = c(as.Date("13-08-2021","%d-%m-%Y"),as.Date("31-10-2021","%d-%m-%Y")),
   y = c(180,200),
   label = c("PAST", "FUTURE")
)

#Time series plots for the next 60 days according to best ARIMA models with 80%–95% CI.
forecast_data_hosp_plot <-fitted_data %>% 
  ggplot()+
  geom_line(aes(x= date, y = data), color = "#5ab4ac")+
  #geom_line(aes(x= date, y = fitted), colour = "red" )+
  geom_line(aes(x= date, y =point_forecast), color ="blue", size = 0.5,
             data = forecast_data )+
  geom_ribbon(aes(x = date, y = point_forecast, ymin = lo_80, ymax = hi_80), 
              data = forecast_data, alpha = 0.3, fill = "green")+
  geom_ribbon(aes(x = date, y = point_forecast, ymin = lo_95, ymax = hi_95), 
              data = forecast_data, alpha = 0.1)+
  geom_vline(aes(xintercept=as.numeric(max(date))),color="#f1a340", linetype="dashed",data = fitted_data)+
  ggtitle("Projection of Hospitalisation") +
  xlab("Month") + 
  ylab("Patient Hospitalised")+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  color_theme()+
  scale_x_date(breaks = "1 month", date_labels = "%b - %y" )+
   geom_text(data=annotation, 
             aes( x=x, y=y, label=label),                  
            color="blue", 
            size=4 )
  
  ggplotly(forecast_data_hosp_plot)
  
```

"Forecast on Deaths (ARIMA Modelling)"

**Data Preparation**

```{r}
#For forecasting, we chose the latest data
trend_death_hb <- trend_hb_daily %>% 
  filter (hb_name == "Scotland") %>% 
  filter(date >="2021-06-01") %>% 
  filter(!(is.na(daily_deaths))) %>% 
  select(date, daily_deaths)

# Convert it into a time series
daily_death_hb_zoo <- zoo(trend_death_hb$daily_deaths, 
           order.by=as.Date(trend_death_hb$date, format='%m/%d/%Y'))

# Convert it into a time series
daily_death_hb_timeseries <-  timeSeries::as.timeSeries(daily_death_hb_zoo)
```

## Step 1 : Visualize the time series

```{r}
original_series_death<-autoplot(daily_death_hb_timeseries, ts.colour = '#5ab4ac')+
  xlab("Month") + 
  ylab("Patient died")+
  #scale_x_date(breaks = "1 month", date_labels = "%b - %y" )+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  ggtitle("Trend on Deaths") +
  color_theme()

ggplotly(original_series_death)
```

## Step 2 : Identification of model : (Finding d:)

Identify whether the time series is stationary / non stationary we can use ADF Augmented Dickey-Fuller test

```{r}
adf_test_death <- adf.test(daily_death_hb_timeseries)
adf_test_death
```

The time series is not stationary since we have a high p-value (p-value must be \< 0.05).
So we apply difference

```{r}
first_diff_death<- diff(daily_death_hb_timeseries)
adf_test1_death <- adf.test(na.omit(first_diff_death))
adf_test1_death
```

Create a dataframe to compare

```{r}
adf_data_death <- data.frame(Data = c("Original", "First-Ordered"),
                       Dickey_Fuller = c(adf_test_death$statistic, adf_test1_death$statistic),
                       p_value = c(adf_test_death$p.value,adf_test1_death$p.value))
adf_data_death
```

Initially the p-value is high which indicates that the Time Series is not stationary.
So we apply difference 1 time.
After the first difference, the p-value \< significance level (0.05) So we can conclude that the difference data are stationary.
So difference (d = 1)

Other method:

```{r}
ndiffs(daily_death_hb_timeseries)
```

Let's plot the First Order Difference Series

Order of first difference

```{r}

first_diff_death<- diff(daily_death_hb_timeseries)
p<- autoplot(first_diff_death, ts.colour = '#5ab4ac') +
  xlab("Month") + 
  ylab("DEATH")+
 # scale_x_date(breaks = "1 month", date_labels = "%b - %y" )+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  ggtitle("First-Order Difference Series") +
  color_theme()

ggplotly(p)
```

## Step 3 Estimate the parameters (Finding p and q)

For our model ARIMA (p,d,q), we found d = 1, the next step is to get the values of p and q, the order of AR and MA part.
Plot ACF and PACF charts to identify q and p respectively.

```{r}
par(mfrow=c(2,2))
acf_death  <- acf1(first_diff_death, col=2:7, lwd=4)
pacf_death <- acf1(first_diff_death,  pacf = TRUE, col=2:7, lwd=4)
```

The ACF and PACF plots tapers to zeo at one point.
So it follows an ARMA model

The PACF non-zero value at lag 3 So the data may follow an AR(3) model

The ACF non - zero value at lag 2 So it can be MA (2)

So we propose three ARMA models for the differenced data: ARMA(p,q) ARMA(3,2), ARMA(3,0) and ARMA(0,2).

That is, for the original time series, we propose three ARIMA models,ARIMA(p,d,q) ARIMA(3,1,2), ARIMA(3,1,0) and ARMA(0,1,2).

## Step 4 Build the ARIMA model

### Manual ARIMA

```{r}
arima_fit_dth_1 = Arima(daily_death_hb_timeseries, order = c(3,1,2))
arima_fit_dth_2 = Arima(daily_death_hb_timeseries, order = c(3,1,0))
arima_fit_dth_3 = Arima(daily_death_hb_timeseries, order = c(0,1,2))
```

```{r}
summary(arima_fit_dth_1)
summary(arima_fit_dth_2)
summary(arima_fit_dth_3)
```

Another way of checking AIC

```{r}
texreg::screenreg(list(arima_fit_dth_1, arima_fit_dth_2, arima_fit_dth_3),
                custom.model.names =c("ARIMA(3,1,2)","ARIMA(3,1,0)","ARIMA(0,1,2)"),
                center = TRUE,
                table = FALSE)
```

Based on this, ARIMA model (3,1,2) seems best.
We can verify the same using automated method

### Automated ARIMA

```{r}
auto_arima_fit_death <- auto.arima(lag(daily_death_hb_timeseries),
                  seasonal=FALSE,
                  stepwise=FALSE,
                  approximation=FALSE,
                  trace = TRUE
                  )
auto_arima_fit_death
```

However Automated ARIMA also confirms that the ARIMA(2,1,3) seems good based on AIC

```{r}
coef_dth<-lmtest::coeftest(auto_arima_fit_death)
coef_dth
```

**Model Selection Criteria :**

ARIMA models with minimum AIC, RMSE and MAPE criteria were chosen as the best models.
Based on Akaike Information Criterion (AIC) above, an ARIMA(2, 1, 3) model seems best.

## Step 5 Check for Diagnostics

Let's plot the diagnostics with the results to make sure the normality and correlation assumptions for the model hold.
If the residuals look like white noise, proceed with forecast and prediction, otherwise repeat the model building.

```{r}
res_dth <-checkresiduals(auto_arima_fit_death, theme = color_theme())
res_dth
```

The ACF plot of the residuals from the ARIMA(2,1,3) model shows that all auto correlations are within the threshold limits except 1, But portmanteau test (Ljung-Box test) shows a higher p-value which means the values are indpendent.
i.e We fail to reject the null hypothesis

Fitting the ARIMA model with the existing data

Let's plot the actuals against the fitted values

**Convert model and time series to dataframe for plotting**

```{r}
daily_death_hb_timeseries_data <- fortify(daily_death_hb_timeseries) %>% 
  clean_names() %>% 
  remove_rownames %>% 
  rename (date = index,
          death = data)%>% 
  mutate(index = seq(1:nrow(daily_death_hb_timeseries)))
  
arima_fit_dth_resid <- ts(daily_death_hb_timeseries[1:nrow(daily_death_hb_timeseries)]) - resid(auto_arima_fit_death)

arima_fit_dth_data <- fortify(arima_fit_dth_resid) %>% 
  clean_names() %>% 
  mutate(data = round(data,2))%>% 
  mutate (data = ifelse(data < 0,0,data))

fit_existing_dth_data <- daily_death_hb_timeseries_data %>% 
  inner_join(arima_fit_dth_data, by = c("index"))


```

**Plotting the series along with the fitted values**

```{r}

fit_existing_dth_plot <- fit_existing_dth_data %>% 
   mutate (data = ifelse(data < 0,0,data)) %>% 
  mutate(date = as.Date(date)) %>% 
  ggplot()+
  aes(x=date, y = death)+
  geom_line(color ="#5ab4ac")+
  geom_line(aes(x= date, y = data), colour = "red" )+
  xlab("Month") + 
  ylab("Deaths reported")+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  scale_x_date(breaks = "1 month", date_labels = "%b - %y" )+
  ggtitle("Fitting the ARIMA model with existing data") +
  color_theme()

ggplotly(fit_existing_dth_plot)

```

## Step 6 Forecast using the model

**Data Preparation :**

```{r}
forecast_dth_model <- forecast(auto_arima_fit_death,level = c(80, 95), h = 30) 

#Convert the model to dataframe for plotting

# Negative values of the CI interval are considered as 0

forecast_dth_model_data <- fortify(forecast_dth_model) %>% 
  clean_names() %>% 
  mutate(data = round(data,2),
         fitted= round(fitted,2))  %>% 
  mutate (lo_80 = ifelse(lo_80 < 0,0,lo_80),
          lo_95 = ifelse(lo_95 < 0,0,lo_95)
          )

forecast_start_date <- as.Date(max(daily_death_hb_timeseries_data$date)+1)
forecast_end_date <- as.Date(forecast_start_date+29)

forecast_dth_data <- forecast_dth_model_data %>% 
  filter(!(is.na(point_forecast))) %>% 
  mutate(date = seq(forecast_start_date,forecast_end_date, by =1)) %>% 
select(-data,-fitted, -index)  

fitted_dth_data <- forecast_dth_model_data %>% 
  filter(!(is.na(data))) %>% 
  inner_join(daily_death_hb_timeseries_data, by = c("index")) %>% 
  mutate(date = as.Date(date)) %>% 
select(date, data, fitted) 

```

**Plotting the Vaccination series plus the forecast and 80 - 95% prediction intervals**

```{r}

annotation <- data.frame(
   x = c(as.Date("13-06-2021","%d-%m-%Y"),as.Date("11-10-2021","%d-%m-%Y")),
   y = c(30,40),
   label = c("PAST", "FUTURE")
)

#Time series plots for the next 60 days according to best ARIMA models with 80%–95% CI.
forecast_data_dth_plot <- fitted_dth_data %>% 
  ggplot()+
  geom_line(aes(x= date, y = data), color = "#5ab4ac")+
  #geom_line(aes(x= date, y = fitted), colour = "red" )+
  geom_line(aes(x= date, y =point_forecast), color ="blue", data = forecast_dth_data )+
  geom_ribbon(aes(x = date, y = point_forecast, ymin = lo_80, ymax = hi_80), 
              data = forecast_dth_data, alpha = 0.3, fill = "green")+
  geom_ribbon(aes(x = date, y = point_forecast, ymin = lo_95, ymax = hi_95), 
              data = forecast_dth_data, alpha = 0.1)+
  geom_vline(aes(xintercept=as.numeric(max(date))),color="#f1a340", linetype="dashed",data = fitted_dth_data)+
  ggtitle("Projection of new Deaths") +
  xlab("Month") + 
  ylab("Death reported")+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  color_theme()+
  scale_x_date(breaks = "1 month", date_labels = "%b - %y" )+
  geom_text(data=annotation, 
            aes( x=x, y=y, label=label),                  
            color="blue", 
            size=4 )
   

ggplotly(forecast_data_dth_plot )
```
